---
output:
  github_document:
    pandoc_args: --webtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

ReferenceData <- read.csv(file.path('Data', 'ReferenceData.csv'))
SampleData <- read.csv(file.path('Data', 'SampleData.csv'))
```

# Code implementation of the subsampling algorithms.

## Introduction

We implemented 3 algorithms, defined “Distance procedure”, the “Probability procedure”, and the “Uniformity procedure” with the aim of producing sub-samples with specific distributional properties starting from an initial, collection of data which is possibly biased in terms of generalizability to a target population. In the following document we present the R code to implement the algorithms. The specific implementation described here is focused on the Italian PPS study on HAI/AMU prevalence in acute care hospitals but can easily translated to different settings.  

The methods try to change the distributional characteristics of a sample by producing a sub-sample whose units are chosen according to some characteristics of interest. Two of these methods, the Distance procedure and the Probability procedure, are aimed at generating a representative sample of a target population, by using reference data at the population level with information on the distribution of relevant characteristics of the observational units. The Uniformity procedure instead generate a sample which is uniform in relation to such characteristics, therefore it does not require population reference data.  

This document and all the relative material is available at https://github.com/AD-Papers-Material/SubsamplingMethods.


## Input data

The algorithms take as input a database with a statistical unit for each row and the characteristics of interest as columns. Furthermore, a column with an unique ID for each unit is useful for post-hoc checks and mandatory for the Distance procedure. Our algorithms also accept the Quality Score ($QS$, cfr. Methods) as additional characteristics, but they can be easily modified to remove such feature.  

The Probability and the Distance procedures also requires population level reference data, with the same structure. If individual level data is not available at the population, simulated data can be generated given access to the joint distribution of the considered characteristics, ensuring that the simulated dataset is large enough to limit random variation.  

Our test case data use acute care hospitals as observational units and hospital size (number of acute care beds) and region of location as characteristics of interest. As a reference, the hospital are also grouped in three hospital size categories; the same categorization is used in the manuscript. We provide simulated sample data for the testing of the procedures at https://github.com/AD-Papers-Material/SubsamplingMethods.


```{r, echo=TRUE}

str(SampleData) # Simulated sample data retaining the real sample characteristics

str(ReferenceData) # Official list of Italian acute hospitals, updated to 2016

```

## General aspects and notation

The procedures at the moment can utilize only categorical characteristics/features of a sample. Therefore continuous characteristics (specifically, hospital size in this case) are discretized in quantiles by the algorithms before use. The number of quantiles to split continuous features into is an input to the algorithms. 
The hospitals are then grouped into *blocks* according to the characteristics of interest: in this study, we used *location/hospital size blocks*, defined by the Italian region ($Region$) and the quantile of number of acute beds ($HSize$) the hospitals fall into. The region and the hospital size quantile allow the definition of a joint discrete probability distribution which is used by the algorithms.
For each block $block_i$ a probability $p_{i} = Pr(hospital|Region, HSize)$ is defined, either given a sample ($p_{i,sample}$) or the whole country ($p_{i,country}$) which indicate the fraction of hospital in a block over the total. All the algorithms are costrained by a parameter $N_{required}$ which defines the size of the final sub-sample. As mentioned above, the algorithms may use the Quality Score $QS$ as further discriminant in the selection; to avoid using the $QS$ without code modification is sufficient to assign the same value (a positive number) to all hospitals. Note that lower $QS$ implies better data quality.

## Unformity procedure

This procedure sub-samples hospitals trying to obtain an equal proportion of hospitals in every block, by iteratively choosing one hospital from each block. This is the general implementation:

* hospitals in the sample are permuted or ordered by $QS$;
* the first hospital (random or with the lower, better, $QS$) is selected;
* all other hospitals belonging to the same block are removed (the block is not available anymore);
* a new hospital is chosen again randomly or by $QS$ for the remaining blocks;
* once one hospital from each blocks has been chosen all the blocks are made available again;
* continue until $N_{required}$ is reached.

For the uniform procedure, we discretized the number of beds only into 4 quantiles since it was less relevant to build a precise discrete probability distribution.

```{r}
subsample.uniform <- function(InputSample, n.required, n.quantiles = 4, use.QS = T){
	library(dplyr)
	library(Hmisc)
 
	Hospitals <- InputSample %>%
		transmute(Code, Region, Beds = Hmisc::cut2(Beds, g = n.quantiles), QS = if (use.QS) QS else 1) # Prepare data by discretizing continuous variables like the number of acute beds and by changing QS to a fixed value if not to be used

	Selected.hospitals <- c()
	Hospitals.temp <- data.frame()

	for (i in 1:n.required) { # Until n.required is reached..
		if (nrow(Hospitals.temp) == 0) { # If no hospitals are already selected store all sample hospital here
			Hospitals.temp <- Hospitals %>%
				filter(!(Code %in% Selected.hospitals)) %>% # Remove already selected hospitals
				sample_frac() %>% # Permute order, useful only if QS is not used
				arrange(QS)
		}

		# Extract the first hospital of the temporary list and add it to the list of selected hospitals
		Extracted.hospital <- Hospitals.temp[1,]
		Selected.hospitals <- c(Selected.hospitals, Extracted.hospital$Code)

		# Remove from the temporary list all hospitals in the same location/size block of the extracted hospital
		Hospitals.temp <- Hospitals.temp %>%
			filter(!(Region %in% Extracted.hospital$Region & Beds %in% Extracted.hospital$Beds))
	}

	InputSample %>% filter(Code %in% Selected.hospitals) # Filter the initial data by the selected hospital codes
}

## Esamples
# Create a subsample of 56 hospitals using the QS
# subsample.uniform(SampleData, n.required = 56)

# The same but this time hospitals are chosen randomly
# subsample.uniform(SampleData, n.required = 56)
```




